---
title: "Reproductive Aging Plots"
author: "Jalen Bynum"
date: "2024-10-30"
output: html_document
---

## Loading packages
```{r}
library(tidyverse)
library(lubridate)
library(dplyr)
library(here)
```

# Cleaning the environment
```{r}
rm(list=ls()) # cleaning the environment first
```

# Loading data and reformatting
Upload breeding data to the environment, format date and calculate time to parturition
```{r cars}
breeding<-read.csv("TS Breeding Book 1 (1).csv")

# Making empty values to NA. NOTE: This makes everything into char variables.
breeding <- breeding %>%
  mutate(across(everything(), na_if, y = ""))

# Checking if the Pup # columns have anything other than NA or numbers
bad_rows <- breeding[!(is.na(breeding$Pups.Birthed) | grepl("^-?\\d+$", breeding$Pups.Birthed)), ]
# All of these are ? entries, and together these 171 entries only have 4 surviving pups. 

#Change number of pups birthed to numeric
breeding$Pups.Birthed<-as.numeric(breeding$Pups.Birthed) 

# Checking if the Pup # columns have anything other than NA or numbers
rm(bad_rows)
bad_rows <- breeding[!(is.na(breeding$Pups.Survived) | grepl("^-?\\d+$", breeding$Pups.Survived)), ]
# All but one of these are ? entries, and together these 6 entries account for 8   pups that were born. I don't think this is a problem given the number of entries we have.

breeding$Pups.Survived<-as.numeric(breeding$Pups.Survived) #Regard number of pups survived as numeric

# Turning dates into the correct format
breeding$IN<-as.Date(breeding$IN,format = "%m/%d/%Y")
breeding$Out<-as.Date(breeding$Out,format = "%m/%d/%Y")
breeding$Birth<-as.Date(breeding$Birth,format = "%m/%d/%Y")


# Defining new variables
breeding$TimeToParturition<-difftime(breeding$Birth,breeding$IN,units="days")
breeding$TimeToParturition <- as.numeric(breeding$TimeToParturition)

breeding$PupSurvival<-breeding$Pups.Survived/breeding$Pups.Birthed #Calculate pup survival

head(breeding)
summary(breeding)
```

For Pups_birthed, we have 177 question mark entries. Together, these only account for 4 surviving offspring. Removing these entries might bias us towards hihger offspring numbers. 


## Data cleaning
Pup numbers look reasonable, and survival goes from 0-1. 

We have some instances of negative parturition, which need to be removed.

The first 2 numbers of each animal ID are the birth year of the animal, so birthyear was derived from animal ID.
```{r}
# Remove entries with negative time to parturition
rm(bad_rows)
bad_rows <- subset(breeding,breeding$TimeToParturition<0)
breeding <-subset(breeding,breeding$TimeToParturition>0)
# Most of these bad rows look like typos, but since we can't say for sure they are removed.

# Sanity check
subset(breeding,breeding$TimeToParturition<0) # Good, nothing shows up

summary(breeding)

```
Now, the date entries - in, out, births - all look reasonable, going from 1996 to 2014.  But there still is a problem with the time to parturition, as it ranges from 3 days to 3699 days. The reported gestation time is 40-52 days, so we should be at least 40 days. I need to have a look at these outliers. 

```{r}
rm(bad_rows)
bad_rows <- subset(breeding, TimeToParturition < 30)

breeding <- subset(breeding, TimeToParturition >3)

rm(bad_rows)
bad_rows <- subset(breeding, TimeToParturition > 200)

breeding <- subset(breeding, TimeToParturition < 3000)

# Sanity check
summary(breeding)
```
There are 11 observations that are below 30 days gestation, which seems unlikely, but 8 of them are supported by the "out" dates as well . I will consider removing these entries. I will remove the entry with 3 days right now, as that is most definitely not correct. 
There are 18 observations above 200 days to parturition. They all make sense given the in/out day, other than the 3699 value, which is clearly some kind of typo. I'm removing this entry. 

## Reformatting some more things to get ages of the parents
The first part of the animal ID is the year they were born.
```{r}

# Modify breeding$year.F for four-digit entries and breeding$year.M for three-digit entries
breeding$year.F <- ifelse(nchar(breeding$F) == 4, substr(breeding$F, 1, 2), substr(breeding$F, 1, 1))
breeding$year.M <- ifelse(nchar(breeding$M) == 4, substr(breeding$M, 1, 2), substr(breeding$M, 1, 1))

# View the updated data frame
head(breeding)
summary(breeding)
```
Ensure that formatting of date is consistent between 1900s and 2000s entries.

```{r}

# Function to convert two-digit or one-digit character years to four-digit years or NA
convert_to_four_digit_year <- function(x) {
  # Check if the entry is numeric and either two-digit or one-digit
  if (grepl("^\\d{1,2}$", x)) {  
    year <- as.numeric(x)
    if (year >= 15 && year <= 99) {
      return(1900 + year)  # Convert two-digit entries to 1900s
    } else if (year >= 1 && year <= 14) {
      return(2000 + year)  # Convert one-digit entries to 2000s
    }
  }
  
  # Return NA for non-numeric entries or entries that don't match the format
  return(NA)
}

# Example usage:
# Assuming 'breeding$year.M' contains your data column
breeding$year.M <- sapply(breeding$year.M, convert_to_four_digit_year)
breeding$year.F<-sapply(breeding$year.F,convert_to_four_digit_year)
# View the updated data frame
head(breeding)
summary(breeding)

```
Ok, the summary shows that we have problem with the birth years, and need to do some more data cleaning here. 

```{r}
rm(bad_rows)
bad_rows <- subset(breeding, year.F < 1990)

breeding <- subset(breeding, year.F > 1989)

rm(bad_rows)
bad_rows <- subset(breeding, year.M < 1990)

breeding <- subset(breeding, year.M > 1987)

# Sanity check
summary(breeding)
```

There are 4 questionable entries for females, and I am going to remove them to be safe.
For males, there are 35 questionable entries, most of them for males with 4 digit IDs that start with 72 or 73. But as these matings are in th 2000s, that's just not correct. I'm keeping one male from 1988, as that makes sense with the females birth data and the mating date, but otherwise, I will remove these, as there might have been something odd with the animal IDs. 


Calculate age using birth year and time of mate pairing (IN).
```{r}
breeding$year.in<-year(breeding$IN) 
breeding$f.age<-as.numeric(difftime(breeding$year.in,breeding$year.F)) #Calculate female age
breeding$m.age<-as.numeric(difftime(breeding$year.in,breeding$year.M)) #Calculate male age

# Sanity check
summary(breeding)
```
Ok, more data cleaning is needed, as we have some negative ages at the time of mate pairings. 

```{r}
rm(bad_rows)
bad_rows <- subset(breeding, f.age < 0)

breeding <- subset(breeding, f.age >= 0)

rm(bad_rows)
bad_rows <- subset(breeding, m.age < 0)

breeding <- subset(breeding, m.age >= 0)

breeding <- subset(breeding, F > 100)
breeding <- subset(breeding, M > 100)

# Sanity check
summary(breeding)
```
These bad entries are mostly due to the animal ID numbers below 100 from the early 2000s, that are not enirely clear. I will remove all the negative entries, and all entries with low animal IDs to be safe. 

A few last checks
```{r}
# Remove rows with NA in TimeToParturition for the entire data frame
breeding <- breeding[!is.na(breeding$TimeToParturition), ]

# Define the quantiles to cut off 5% on upper end of time to parturition
upper_bound <- quantile(breeding$TimeToParturition, 0.95)

# Filter the data to remove outliers
cleaned.data <- breeding[ breeding$TimeToParturition <= upper_bound, ]
outliers<-breeding[breeding$TimeToParturition>upper_bound,]
```
If we are removing outliers, maybe we should do that at both ends of the distribution...


## Making some frequency tables and some basic analysis
```{r}

F.age.table<-table(cleaned.data$f.age)
M.age.table<-table(cleaned.data$m.age)

parturition.table<-table(cleaned.data$TimeToParturition) # not helpful; too many bins

# Calculate group means
vars_to_summarise <- c("Pups.Birthed", "Pups.Survived", "PupSurvival", "TimeToParturition")

summary_table.f <- breeding %>%
  group_by(f.age) %>%
  summarise(
    across(
      all_of(vars_to_summarise),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        se = ~ sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))),
        n = ~ sum(!is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    )
  )

summary_table.f

summary_table.m <- breeding %>%
  group_by(m.age) %>%
  summarise(
    across(
      all_of(vars_to_summarise),
      list(
        mean = ~ mean(.x, na.rm = TRUE),
        se = ~ sd(.x, na.rm = TRUE) / sqrt(sum(!is.na(.x))),
        n = ~ sum(!is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    )
  )

summary_table.m

# Non-parametric tests for the effect of age
kruskal.test(Pups.Birthed ~ f.age, data = cleaned.data) # Sign
kruskal.test(Pups.Birthed ~ m.age, data = cleaned.data) # Sign

kruskal.test(Pups.Survived ~ f.age, data = cleaned.data) # Sign
kruskal.test(Pups.Survived ~ m.age, data = cleaned.data) # ns

kruskal.test(PupSurvival ~ f.age, data = cleaned.data) # Sign
kruskal.test(PupSurvival ~ m.age, data = cleaned.data) # ns

kruskal.test(TimeToParturition ~ f.age, data = cleaned.data) # Sign
kruskal.test(TimeToParturition ~ m.age, data = cleaned.data) # Sign

# Following up with Dunn's tests
library(dunn.test)
attach(cleaned.data)
dunn.test(Pups.Birthed, f.age, kw=FALSE, method="bonferroni")
dunn.test(Pups.Birthed, m.age, kw=FALSE, method="bonferroni")

dunn.test(Pups.Survived, f.age, kw=FALSE, method="bonferroni")
dunn.test(Pups.Survived, m.age, kw=FALSE, method="bonferroni")

dunn.test(PupSurvival, f.age, kw=FALSE, method="bonferroni")
dunn.test(PupSurvival, m.age, kw=FALSE, method="bonferroni")

dunn.test(TimeToParturition, f.age, kw=FALSE, method="bonferroni")
dunn.test(TimeToParturition, m.age, kw=FALSE, method="bonferroni")
detach(cleaned.data)
```



```{r}
# Keeping for later, just in case
#data$f.age<-na.omit(data$f.age)
#data$m.age<-na.omit(data$m.age)
#data$Pups.Survived<-na.omit(data$Pups.Survived)

```


## Testing some basic linear models
```{r}
data_long <- cleaned.data %>%
  pivot_longer(cols = c(f.age, m.age), 
               names_to = "Age_Type", 
               values_to = "Age")

# Ensure Age_Type is a factor
data_long$Age_Type <- as.factor(data_long$Age_Type)

# Correlation between male and female age in mate pair
cor.test(cleaned.data$f.age, cleaned.data$m.age)

# Create linear model - basic linear model for time
model_f_age <- lm(TimeToParturition ~ f.age, data = cleaned.data)
summary(model_f_age)
model_m_age <- lm(TimeToParturition ~ m.age, data = cleaned.data)
summary(model_m_age)
model_fm_age <- lm(TimeToParturition ~ f.age * m.age, data = cleaned.data)
summary(model_fm_age)

# here, the dependent variable is not continuous, so need to use logistic regression
rm(model_f_age, model_m_age, model_fm_age)
model_f_age <- glm(Pups.Birthed ~ f.age, family = poisson(link = "log"), data = cleaned.data) # Significant
summary(model_f_age)
model_m_age <- glm(Pups.Birthed ~ m.age, family = poisson(link = "log"), data = cleaned.data)
summary(model_m_age)
model_fm_age <- glm(Pups.Birthed ~ f.age * m.age, family = poisson(link = "log"), data = cleaned.data)
summary(model_fm_age)

rm(model_f_age, model_m_age, model_fm_age)
model_fm_age <- glm(Pups.Survived ~ f.age * m.age, family = poisson(link = "log"), data = cleaned.data)
summary(model_fm_age)

# This is continous data from 0 to 1, and needs a beta regression
rm(model_f_age, model_m_age, model_fm_age)
library (betareg)
# Scaling PupSurvival so that it doesn't includes 0 and 1's as those would cause errors.
n <- nrow(cleaned.data)
cleaned.data$PupSurvival2 <- (cleaned.data$PupSurvival * (n - 1) + 0.5) / n

model_fm_age <- betareg(PupSurvival2 ~ f.age * m.age, data = cleaned.data)
summary(model_fm_age)


```


```{r}
# Create the plot with boxplots, points, and goodness-of-fit lines
PupSurvivalPlot<-ggplot(data_long, aes(x = as.factor(Age), y = Pups.Survived)) +
  geom_boxplot(aes(fill = Age_Type), position = position_dodge(0.75), alpha = 0.6) + 
  geom_point(aes(color = Age_Type), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75), alpha = 0.6) +
  labs(title = "Boxplots of Time to Parturition with respect to Parental Age",
       x = "Time",
       y = "Pups Survived",
       fill = "Age Type",
       color = "Age Type") +
  theme_minimal() +
  scale_fill_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_color_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_y_continuous(limits = c(0, 4))

PupSurvivalPlot
#Model the relationship between parental age and number of pups survived
pup.survival.model <- lm(Pups.Survived ~ m.age * f.age, data = cleaned.data)
summary(pup.survival.model)
```
## Plot the time to parturition by parental age.
```{r}

data_long <- cleaned.data %>%
  pivot_longer(cols = c(f.age, m.age), 
               names_to = "Age_Type", 
               values_to = "Age")
# Female and Male ages were considered age values
# Ensure Age_Type is a factor
data_long$Age_Type <- as.factor(data_long$Age_Type)


# Create the plot with boxplots, points, and goodness-of-fit lines
ParturitionPlot<-ggplot(data_long, aes(x = as.factor(Age), y = TimeToParturition)) +
  geom_boxplot(aes(fill = Age_Type), position = position_dodge(0.75), alpha = 0.6) + 
  geom_point(aes(color = Age_Type), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75), alpha = 0.6) +
  labs(title = "Boxplots of Conception Time across Ages",
       x = "Time",
       y = "Time to Parturition (days)",
       fill = "Age Type",
       color = "Age Type") +
  theme_minimal() +
  scale_fill_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_color_manual(values = c("f.age" = "red", "m.age" = "blue"))
ParturitionPlot

# Saving a file
pdf(file = "TimeToBirth.pdf", width = 8, height = 5)

ParturitionPlot

dev.off()
```

## Plot the number of Pups by parental age

```{r}
# Create the plot with boxplots & points
PupsBornPlot<-ggplot(data_long, aes(x = as.factor(Age), y = Pups.Birthed)) +
  geom_boxplot(aes(fill = Age_Type), position = position_dodge(0.75), alpha = 0.6) + 
  geom_point(aes(color = Age_Type), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75), alpha = 0.6) +
  labs(title = "Boxplots of pups birthed across Ages",
       x = "Time",
       y = "Number of Pups birthed",
       fill = "Age Type",
       color = "Age Type") +
  theme_minimal() +
  scale_fill_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_color_manual(values = c("f.age" = "red", "m.age" = "blue"))
PupsBornPlot

violin_birthed <- ggplot(data_long, aes(fill=Age_Type, y=Pups.Birthed, x=as.factor(Age))) + 
    geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
    labs(title = "Boxplots of pups birthed across Ages",
       x = "Time",
       y = "Number of Pups birthed",
       fill = "Age Type",
       color = "Age Type") +
  theme_minimal() +
  scale_fill_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_color_manual(values = c("f.age" = "red", "m.age" = "blue"))

violin_birthed

# Saving the plot
pdf(file = "PupsBirthed.pdf", width = 8, height = 5)
violin_birthed
dev.off()

violin_survived <- ggplot(data_long, aes(fill=Age_Type, y=Pups.Survived, x=as.factor(Age))) + 
    geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent") +
    labs(title = "Boxplots of pups birthed across Ages",
       x = "Time",
       y = "Number of Pups survived",
       fill = "Age Type",
       color = "Age Type") +
  theme_minimal() +
  scale_fill_manual(values = c("f.age" = "red", "m.age" = "blue")) +
  scale_color_manual(values = c("f.age" = "red", "m.age" = "blue"))

violin_survived

# Saving the plot
pdf(file = "PupsSurvived.pdf", width = 8, height = 5)
violin_survived
dev.off()

```
Looks bad - can we do violin plot instead...