---
title: "TS_manuscript_Longevity"
author: "Jalen Bynum"
date: "2024-10-31"
output: html_document
---

# Before getting started
Install Packages necessary to complete the analysis:
ggfortify
tidyverse
survival
survminer
readxl
lubridate
hrbrthemes
ggridges
devtools
BaSTA2.0
writexl
here


# Loading required packages
```{r}
library(ggfortify)
library(tidyverse)
library(survival)
library(survminer)
library(readxl)
library(lubridate)
library(hrbrthemes)
library(ggridges)
library(devtools)
# library(BaSTA2.0) # version not available for this version of R fix later
library(writexl)
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)
library(flexsurv)
library(here)
```

# Load data set.

```{r}
rm(list=ls()) # cleaning the environment first
dat<-read.csv("TS_final_data.csv")

```

# Reformatting and cleaning the data
Matrices were created to format entries of sex and define NA values.
```{r}
cut <- c("N/A", "?", " ","")
male<- c("M ","m"," M")
female<-"f"
```

Entries in the Sex column were reformatted to either "male" or "female". Incomplete or unclear entries in the Sex and DOB columns were marked as NA.
```{r}
dat[dat[,3] %in% cut, 3] <-NA # fixing NAs in DOB

dat[dat[,2] %in% male,2] <-"M"  # change "male" to "M"
dat[dat[,8]=="N ",8]<- "N" # making " N" and "N" consistent
dat[dat[,2] %in% female,2]<-"F"  # change "female to "F"
dat[dat[,2] %in% cut, 2] <-NA # fixing NAs in sex
dat[dat[,7] %in% cut, 7] <-NA # fixing NAs in death date

dat$Sex <- as.factor(dat$Sex) # making sex a factor

summary(dat)  #checking how things look
```

# Checking duplicate entries.

```{r}
# Check for duplicate entries
flagged <- dat[duplicated(dat$Animal_ID), ]
summary(flagged)
```

No duplicates left. 

Incomplete cases of DOB or Animal ID were removed from the data set.
```{r}
step.2<-dat[complete.cases(dat$DOB),]
step.3<-step.2[complete.cases(step.2$Animal_ID),]
summary(step.3)
```

Date is formatted and undefined sex entries are removed
```{r}
step.3$DOB<-as.Date(step.3$DOB, "%m/%d/%y") # if not in the right format, returns NA
step.3 <-step.3[complete.cases(step.3$DOB),] # keep only the entries that have a DOB

step.3$Death_date<-as.Date(step.3$Death_date, "%m/%d/%y") # if not in the righ formati, retunrs NA

# Removing entries without sex information
step.3a<-step.3[complete.cases(step.3$Sex),] # keep only entries with sex specified
# Sanity check to make sure we have two levels for sex.
unique(step.3a$Sex)

# Checking the sex ration, with #'s of F and M from summary
chisq.test(x=c(687, 566), p=c(0.5, 0.5))
# Sex ratio is biased towards females.
summary(step.3a)
```
Calculate lifespan, remove negative and 0 day lifespans. 
For living animals, we will assign a "death date" of 2023-08-01. These animals, as well as the animals that were euthanized, will be coded as "right censored" as we don't know their natural death date. 
Note: We had one additional death added later to the dataset (2024-10-20).
```{r}
liv.censor.date<-as.Date("2023-08-01")
liv.censor.date<-as.Date(liv.censor.date)
step.3a[step.3a[,8]=="L",7]<- liv.censor.date
liv<-subset(step.3a,SAC=="L")
liv
step.3a$lifespan<-difftime(step.3a$Death_date, step.3a$DOB, units = "days")
# Sanity check
#hist(as.numeric(step.3a$lifespan, units = "days"))
# shows we have some negative numbers

 negatives<-subset(step.3a, (lifespan < 0)) # look at entries with neg numbers
# Those entries were checked and the issue can't be resolved.

# Remove enties with negative lifespan
step.4 <- subset(step.3a, lifespan >= 0 | is.na(lifespan))

# Remove entries that don't have lifespans
step.4 <-filter(step.4,!is.na(lifespan))

# Remove entries without cause of death
step.4<-subset(step.4,!SAC=="") # Remove all animals with unknown deaths

summary(step.4)

# Checking the sex ration, with #'s of F and M from summary
chisq.test(x=c(533, 436), p=c(0.5, 0.5))
# Sex ratio is biased towards females. 

```


Make a table to see the number of animals in these groups.
```{r}
step.4[step.4[,8] == "L",8] <-"Living"
step.4[step.4[,8] == "Y",8] <-"Euthanized"
step.4[step.4[,8] == "N",8] <-"Natural Death"
step.4[step.4[,8] == "A",8] <-"Right censored" #Animals without a death date, but a last seen alive date
table(step.4$Sex,step.4$SAC)

```

## Checking the sex ration in these groups
```{r}
chisq.test(c(F = 310, M = 263)) # Sign.
chisq.test(c(F = 89, M = 75))
chisq.test(c(F = 109, M = 75)) # Sign.
chisq.test(c(F = 25, M = 23))
```
We have more females than males overall and among the euthanized and natural death animals. 

##################################################################################################
## Data exploration
##################################################################################################

Currently, lifespan is encoded as a "difftime" object, not as an actual number. To be able to do the various analysis, we need to change that to numeric. It will be in days.

```{r}
step.4$lifespan <- as.numeric(step.4$lifespan)
```


Recoding things for the lifespan analysis 
```{r}

data.fit<-step.4%>%mutate(SAC=ifelse(!SAC=="Natural Death",0,1))%>%mutate_if(is.character,as_factor) #Recodes cause of deaths as 0/1 for censoring

```


# Distribution of births
Ridge Plot of DOB distribution according to Sex
Females in Red
Males in Blue
```{r}
dis.DOB.sex<-data.fit%>%ggplot( aes(x=DOB,y=Sex,fill=Sex)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#D81B60", "#1E88E5","#FFC107","#004D40")) +theme_ipsum() +labs(fill="Sex")+ggtitle("Birthdate Distribution by Sex")
print(dis.DOB.sex)
```

Are there any differences in DOB between the sexes? 
```{r}
kruskal.test(as.numeric(data.fit$DOB) ~ data.fit$Sex)
```

# Distribution of death deates
Ridge plot of death date distribution according to sex

```{r}
dis.Death.sex<-subset(step.4, (SAC != "Living" & SAC != "Right censored")) %>%ggplot( aes(x=Death_date,y=Sex,fill=Sex)) +ggtitle("Bin size = 3") +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#D81B60", "#1E88E5","#004D40")) +theme_ipsum() +labs(fill="Sex")+ggtitle("Death Date Distribution by Sex")
print(dis.Death.sex)
```


Are there sex differences in the death dates?
```{r}
x <- step.4 %>% subset((SAC != "Living" & SAC != "Right censored"))
kruskal.test(as.numeric(x$Death_date) ~ x$Sex)
kruskal.test(Death_date ~ Sex, data = subset(step.4, (SAC != "Living" & SAC != "Right censored"))) # Same result
rm(x)
```


Reformat cause of death for readability.
Make excel file of death contingency table

```{r}

mort.table<-table(data.fit$lifespan,data.fit$Sex)
if (!is.data.frame(mort.table)) {
  mort.table <- as.data.frame(mort.table)
}
mort_by_sex<-write_xlsx(mort.table,path="mort_by_sex.xlsx")
```

Make subsections to view cause of death
```{r}
euth<-subset(step.4,step.4$SAC=="Euthanized")
nd<-subset(step.4,step.4$SAC=="Natural Death")
rc<-subset(step.4,step.4$SAC=="Right censored")
living<-subset(step.4,step.4$SAC=="Living")
```

Ridge Plot of lifespan distribution according to Sex
Females in Red
Males in Blue
```{r}
dis.Sex<-data.fit %>% subset((SAC != "Living" & SAC != "Right censored")) %>%ggplot( aes(x=lifespan/365.25,y=Sex,fill=Sex)) +geom_density_ridges() +
  theme_ridges() +
  theme(legend.position = "none") +scale_fill_manual(values=c("#D81B60", "#1E88E5","#FFC107","#004D40")) +theme_ipsum() +labs(fill="Sex", x = "Lifespan (years)")+ggtitle("Lifespan Distribution by Sex ")
print(dis.Sex)
```
Testing for overall lifespan differences by sex irrespective of cause of death
```{r}
kruskal.test(lifespan ~ Sex,data=subset(step.4, (SAC != "Living" & SAC != "Right censored")))

step.4 %>% subset(SAC != "Living" & SAC != "Right censored") %>%
  group_by(Sex) %>%
  summarize(mean = mean(lifespan), median = median(lifespan)) 

# and with just natural death animals
kruskal.test(lifespan ~ Sex,data=subset(data.fit, (SAC == 1)))

data.fit %>% subset(SAC ==1) %>%
  group_by(Sex) %>%
  summarize(mean = mean(lifespan), median = median(lifespan)) 
```
## Ridge Plot of Lifespan distribution according to cause of death
Euthanized animals in green
natural death animals in blue
unknown animals in orange

```{r}
dis.Sac<-step.4%>%filter(!SAC %in% c("Living", "Right censored"))%>%ggplot( aes(x=lifespan/365.25,y=SAC,fill=SAC)) +geom_density_ridges() +
  theme_ridges() +
  theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +labs(fill="Cause Of Death", x = "Lifespan (years)")+ggtitle("Lifespan Distribution by Sac ")
print(dis.Sac)


```
Testing for lifespan differences between euthanized and natural death animals
```{r}
x <- step.4%>%filter(!SAC %in% c("Living", "Right censored"))
kruskal.test(x$lifespan ~ x$Sex) #checking for sex differences
kruskal.test(x$lifespan ~ x$SAC) #checking for differences by death type
rm(x)

# Getting some averages
x <- step.4 %>% subset (SAC == "Natural Death")
mean(x$lifespan, na.rm = TRUE)
rm(x)

x <- step.4 %>% subset (SAC == "Euthanized")
mean(x$lifespan, na.rm = TRUE)
rm(x)

x <- step.4 %>% filter(!SAC %in% c("Living", "Right censored")) %>% subset (Sex == "F")
mean(x$lifespan, na.rm = TRUE)
rm(x)

x <- step.4 %>% filter(!SAC %in% c("Living", "Right censored")) %>% subset (Sex == "M")
mean(x$lifespan, na.rm = TRUE)
rm(x)

step.4 %>% filter(SAC == "Natural Death") %>%
  filter (Sex == "F") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 

step.4 %>% filter(SAC == "Natural Death") %>%
  filter (Sex == "M") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 
      
step.4 %>% filter(SAC == "Living") %>%
  filter (Sex == "F") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 
            
step.4 %>% filter(SAC == "Living") %>%
  filter (Sex == "M") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 

step.4 %>% filter(SAC == "Euthanized") %>%
  filter (Sex == "F") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 
            
step.4 %>% filter(SAC == "Euthanized") %>%
  filter (Sex == "M") %>%
  summarise(mean_lifespan = mean(lifespan, na.rm = TRUE), 
            median_lifespan = median(lifespan, na.rm = TRUE)) 
```


## Ridge plot of lifespan distribution according to sex for natural death animals

```{r}
dis.nd<-nd%>%ggplot( aes(x=lifespan/365.25, y=Sex,fill=Sex, jittered_points = TRUE)) +geom_density_ridges() +
  theme_ridges() +
  theme(legend.position = "none") +scale_fill_manual(values=c("#D81B60", "#1E88E5","#FFC107","#004D40")) +theme_ipsum() +labs(fill="Sex", x = "Lifespan (years)")+ggtitle("Lifespan Distribution by Sex of Natural Death Animals")
print(dis.nd)
```

Testing for sex differences in natural death animals
```{r}
kruskal.test(nd$lifespan ~ nd$Sex) #checking for sex differences
```


## Ridge plot of death date distribution according to cause of death

```{r}
death.fit<-step.4 %>% filter(!SAC %in% c("Living", "Right censored"))
summary(death.fit)
dis.Death.sac<-death.fit%>%ggplot( aes(x=Death_date,y=SAC,fill=SAC)) +ggtitle("Bin size = 3") +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +ggtitle("Death Date Distribution by Sac")
print(dis.Death.sac)
```

Checking for differences by sex and death type
```{r}
x <- death.fit
kruskal.test(as.numeric(x$Death_date) ~ x$Sex) #checking for sex differences
kruskal.test(as.numeric(x$Death_date) ~ x$SAC) #checking for differences by death type
rm(x)
```
## Ridge plot of date of birth distribution according to cause of death

```{r}
dis.DOB.sac<-step.4%>%ggplot( aes(x=DOB,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +labs(fill="SAC")+ggtitle("Birthdate Distribution by Sac")
print(dis.DOB.sac)
```


Create male and female subsections for main data set and for dead animals dataset.
```{r}
# step.4 includes all the cleaned data, even individuals with unknown cause of death.
m.cage<-subset(step.4,Sex== "M",)
f.cage<-subset(step.4,Sex=="F",)
```

```{r}
# death.fit includes all dead animals, and excludes data from animals that are alive
m.death<-subset(death.fit,Sex=="M")
f.death<-subset(death.fit,Sex=="F")
```


## Lifespan distribution of females according to cause of death
```{r}
dis.lifespan.F<-step.4%>%filter(Sex=="F")%>%filter(lifespan>0)%>%ggplot( aes(x=lifespan/365.25,y=SAC,fill=SAC)) +geom_density_ridges() +
  theme_ridges() +
  theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +ggtitle("Bin size = 3")+labs(fill="Sac", x = "Lifespan (years)")+ggtitle("Lifespan Distribution of Females by Sac")
print(dis.lifespan.F)

kruskal.test(lifespan ~ SAC, data = f.cage)
```

## Ridge plot of DOB distribution for females according to cause of death
```{r}
dis.DOB.F.L<-f.cage%>%ggplot( aes(x=DOB,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +labs(fill="Sex")+ggtitle("Birthdate Distribution of Females by Sac")
print(dis.DOB.F.L)

kruskal.test(as.numeric(f.cage$DOB) ~ f.cage$SAC)
```

  
## Ridge plot of death date distribution for females according to cause of death
```{r}
dis.Death.F<-f.death%>%filter(lifespan>0)%>%ggplot( aes(x=Death_date,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +ggtitle("Death Date Distribution of Females by Sac")
print(dis.Death.F)

kruskal.test(as.numeric(f.death$Death_date) ~ f.death$SAC)
```


## Rigde plot of Lifespan distribution for males according to cause of death

```{r}
dis.lifespan.M.L<-m.cage%>%ggplot( aes(x=lifespan/365.25,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum()+labs(fill="Sac", x = "Lifespan (years)")+ggtitle("Lifespan Distribution of Males by Sac")
print(dis.lifespan.M.L)

kruskal.test(lifespan ~ SAC, data = m.cage)
```

## Ridgeplot of date of birth distribution for males according to cause of death

```{r}
dis.DOB.M.L<-m.cage%>%ggplot( aes(x=DOB,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +labs(fill="SAC")+ggtitle("Birthdate Distribution of Males by Sac")
print(dis.DOB.M.L)

kruskal.test(as.numeric(m.cage$DOB) ~ m.cage$SAC)
```


## Ridgeplot of death date distribution for males according to cause of death

```{r}
dis.Death.M<-m.death%>%filter(lifespan>0)%>%ggplot( aes(x=Death_date,y=SAC,fill=SAC)) +geom_density_ridges() +   theme_ridges() +   theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#CA6024","#F05151","#258EF8","#F825C5")) +theme_ipsum() +ggtitle("Death Date Distribution of Males by Sac")
print(dis.Death.M)

kruskal.test(as.numeric(m.death$Death_date) ~ m.death$SAC)
```

## One more Ridge plot to save for the paper
## Ridge Plot of Lifespan distrbution according to cause of death
Euthanized animals in green
natural death animals in blue
living animals in orange

```{r}
step.4$SAC <- factor(step.4$SAC, levels=c('Euthanized', 'Natural Death', 'Living', 'Right censored'))
plot<-step.4%>%filter(!SAC %in% c("Unknown"))%>%ggplot( aes(x=lifespan/365.25,y=SAC,fill=SAC)) +geom_density_ridges(rel_min_height = 0.01) +
  # xlim (0, 12) +
  theme_ridges() +
  theme(legend.position = "none") +scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) +labs(fill="Cause Of Death", x = "Lifespan (years)")+ggtitle("Lifespan Distribution by Sac ")
print(plot)


```
NOTE from 05/05/2026
After feedback from the co-authors, we decided that this plot needs to be modified, as right now, the smoothing shows negative lifespans. I will generate a few options to see what people like best. 


```{r}
step.4$SAC <- factor(step.4$SAC, levels=c('Euthanized', 'Natural Death', 'Living', 'Right censored'))
plot1<-step.4%>%filter(!SAC %in% c("Right censored"))%>%
  ggplot( aes(x=lifespan/365.25, y=SAC,fill=SAC, height = after_stat(density))) +
  geom_density_ridges(stat = "density") +
  theme_ridges() +
  theme(legend.position = "none") +
    scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) + 
    labs(fill="Cause Of Death", x = "Lifespan (years)") +
    ggtitle("Lifespan Distribution by Sac")
  
print(plot1)

step.4$SAC <- factor(step.4$SAC, levels=c('Euthanized', 'Natural Death', 'Living', 'Right censored'))
plot2<-step.4%>%filter(!SAC %in% c("Right censored"))%>%
  ggplot( aes(x=lifespan/365.25, y=SAC,fill=SAC, height = after_stat(density))) +
  geom_density_ridges(stat = "binline", scale = 0.95) +
  theme_ridges() +
  theme(legend.position = "none") +
    scale_fill_manual(values=c("#69b3a2", "#404080","#CA6024","#F05151","#258EF8","#F825C5")) + 
    labs(fill="Cause Of Death", x = "Lifespan (years)") +
    ggtitle("Lifespan Distribution by Sac")
  
print(plot2)

```
Sent to co-authors on 5/5/2026. We decided to use plot2 for the paper. 


```{r}
#T.belangeri ridge plot by cause of death
pdf(file = "Ridge.plot2.SAC.pdf", width = 8, height = 5)

plot2

dev.off()

```

NOTE: Theme ipsum used in all these ridge plots means it can't be exported to pdf due to an issue with the fonts. Removing the theme fixes that. 

We also need some information about how many animals we have that are in various age classes, particularly to see how many we have in the older age classes. This bit of code will give us a summary table. 

```{r}
table(step.4$Sex,step.4$SAC) # checking how many animals per "death type"

# Create bins and count
result <- step.4 %>%
  filter (!is.na(lifespan)) %>%
  mutate(bin = cut(lifespan, breaks = c(0,365.25, 2*365.25, 3*365.25, 4*365.25, 5*365.25, 6*365.25, 7*365.25))) %>%
  group_by(bin) %>%
  summarise(count = n())

print(result)

step.4 %>% 
  mutate (SAC = as.factor(SAC)) %>% 
  filter (lifespan > 365.25*7) %>% 
  summary()

step.4 %>% 
  mutate (SAC = as.factor(SAC)) %>% 
  filter (lifespan > 365.25*6) %>% 
  summary()

step.4 %>% 
  mutate (SAC = as.factor(SAC)) %>% 
  filter (lifespan > 365.25*5) %>% 
  summary()


```


## Preparing for survival analysis
Define natural death as uncensored and all other entries as censored
Natural death encoded as "1", euthanized or living encoded as "0" in SAC column
```{r}
# Assuming 'status' is the variable indicating if an animal is alive, with "Alive" as the label
# dataframe created in line 206
summary(data.fit)

```

## Survival analysis
Kaplan Meier plot of survival probability for males and females.
Natural Death animals are considered uncensored while living and euthanized entries are right censored
Male survival probability is in red and female survival probability is in blue
```{r}

surv <- Surv(time = data.fit$lifespan/365.25, event = data.fit$SAC, type = "right")
fitsex <- survfit(formula= surv ~ Sex, data = data.fit) # uses default Kaplan-Meier
summary(fitsex)
kap.sex<-ggsurvplot(
  fit = fitsex, 
  data = data.fit,
  pval = TRUE, 
  surv.median.line = "hv",
  xlab = "Survival time (years)", 
  ylab = "Survival probability", 
  title = "Kaplan Meier's Plot by Sex"
  )
print(kap.sex)

# Exporting the file
pdf(file = "Kaplan_Meier_all.pdf", width = 8, height = 5)

kap.sex

dev.off()
```

## Additional test for lifespan differences
```{r}
# Comparing the lifespan curves
log_rank_test <- survdiff(Surv(lifespan, SAC) ~ Sex, data = data.fit)
log_rank_test

# Comparing median lifespan
wilcox.test(lifespan ~ Sex, alternative = "greater", data = data.fit)
```


```{r}

# Assuming `lifespan` is your survival time variable and `Sex` is coded as 'F' and 'M'
surv_object <- Surv(time = data.fit$lifespan/365.25, event = data.fit$SAC)  # Replace 'event' with your event indicator

# Fit Kaplan-Meier survival curves, stratified by Sex
km_fit <- survfit(surv_object ~ Sex, data = data.fit)

# Plot the Kaplan-Meier curves
ggsurvplot(km_fit,
           data = data.fit,
           pval = TRUE,                  # Show p-value for log-rank test
           conf.int = TRUE,              # Add confidence intervals
           risk.table = TRUE,            # Add risk table below the plot
           legend.labs = c("Female", "Male"), # Adjust labels
           ggtheme = theme_minimal())    # Set theme

```



```{r}
# Fit Kaplan-Meier survival curves for males and females
km_fit <- survfit(Surv(lifespan/365.25, SAC) ~ Sex, data = data.fit)

# Define quantiles of interest (e.g., 50% and 25%)
quantile_probs <- c(0.25, 0.5,0.75)

# Extract survival times corresponding to the quantiles for each group
surv_times <- quantile(km_fit, probs = quantile_probs)

# Display survival times at 50% and 25% survival for each group
print(surv_times)

# Calculate the absolute difference in survival times between groups at these quantiles
# Example: Difference in survival probability at specific time points (when 25% or 50% are left)
female_surv <- km_fit$surv[km_fit$strata == "Sex=F"]
male_surv <- km_fit$surv[km_fit$strata == "Sex=M"]

# Calculate differences at median survival and 25% survival points
median_diff <- abs(female_surv - male_surv)
quantile_diff <- abs(female_surv[quantile_probs == 0.5] - male_surv[quantile_probs == 0.5])

# Display differences
median_diff
quantile_diff

```
## Looking only at Natural deaths
```{r}
rm(surv)
rm(fitsex)
rm(kap.sex)
rm(log_rank_test)
nd_data <-data.fit %>% filter(SAC == "1")
surv <- Surv(time = nd_data$lifespan/365.25, event = nd_data$SAC, type = "right")
fitsex <- survfit(formula= surv ~ Sex, data = nd_data) # uses default Kaplan-Meier
summary(fitsex)
kap.sex<-ggsurvplot(
  fit = fitsex, 
  data = nd_data,
  pval = TRUE, 
  surv.median.line = "hv",
  xlab = "Survival time (years)", 
  ylab = "Survival probability", 
  title = "Kaplan Meier's Plot by Sex"
  )
print(kap.sex)

# Exporting the file
pdf(file = "Kaplan_Meier_natural.pdf", width = 8, height = 5)

kap.sex

dev.off()

# Comparing the lifespan curves
log_rank_test <- survdiff(Surv(lifespan, SAC) ~ Sex, data = nd_data)
log_rank_test

# Comparing median lifespan
wilcox.test(lifespan ~ Sex, alternative = "greater", data = nd_data)
```

## And looking at animals that are natural deaths and alive
```{r}
rm(surv)
rm(fitsex)
rm(kap.sex)
rm(log_rank_test)
nda_data <-step.4 %>% filter(SAC != "Euthanized") %>%mutate(SAC=ifelse(!SAC=="Natural Death",0,1))%>%mutate_if(is.character,as_factor)
summary(nda_data)
surv <- Surv(time = nda_data$lifespan/365.25, event = nda_data$SAC, type = "right")
fitsex <- survfit(formula= surv ~ Sex, data = nda_data) # uses default Kaplan-Meier
summary(fitsex)
kap.sex<-ggsurvplot(
  fit = fitsex, 
  data = nda_data,
  pval = TRUE, 
  surv.median.line = "hv",
  xlab = "Survival time (years)", 
  ylab = "Survival probability", 
  title = "Kaplan Meier's Plot by Sex"
  )
print(kap.sex)

# Exporting the file
pdf(file = "Kaplan_Meier_excluding_euth.pdf", width = 8, height = 5)

kap.sex

dev.off()

# Comparing the lifespan curves
log_rank_test <- survdiff(Surv(lifespan, SAC) ~ Sex, data = nda_data)
log_rank_test

# Comparing median lifespan
wilcox.test(lifespan ~ Sex, alternative = "greater", data = nda_data)
```
# Looking at maximum lifespan differences
Fisher's test for maximum lifespan (Wang 2004)
```{r}
# compute 90th percentile cutoff from pooled distribution
cutoff <- quantile(data.fit$lifespan, 0.9, na.rm = TRUE)


  # classify long-lived vs not
  df <- data.fit %>%
    mutate(
      top10 = ifelse(lifespan >= cutoff, "yes", "no"),
      group = Sex
    )

  # build contingency table
  tab <- table(df$group, df$top10)

  # run fisher test
  ft <- fisher.test(tab)
  ft

  
# Doing the same with the natural death only data
rm(cutoff, df, tab, ft)

cutoff <- quantile(nd_data$lifespan, 0.9, na.rm = TRUE)

  df <- nd_data %>%
    mutate(
      top10 = ifelse(lifespan >= cutoff, "yes", "no"),
      group = Sex
    )

  tab <- table(df$group, df$top10)

  ft <- fisher.test(tab)
  ft
  
# And with the alive and natural death animals
rm(cutoff, df, tab, ft)

cutoff <- quantile(nda_data$lifespan, 0.9, na.rm = TRUE)

  df <- nda_data %>%
    mutate(
      top10 = ifelse(lifespan >= cutoff, "yes", "no"),
      group = Sex
    )

  tab <- table(df$group, df$top10)

  ft <- fisher.test(tab)
  ft  
```
 

# Mortality analysis
I'll make the mortality curves and test for differences. 

```{r}
# 0 Recode Status to numeric (Dead = 1, others = 0)
# This first attempt only has natural death data.
summary(nd_data)

# 1 Function: actuarial/demography life table at event times
make_demography_lifetable <- function(subdf, time_col = "lifespan", status_col = "SAC") {
  # keep only rows with non-missing Death_Day/status
  subdf <- subdf %>% filter(!is.na(.data[[time_col]]), !is.na(.data[[status_col]])) %>%
    arrange(.data[[time_col]])
  
  # fit KM for this subgroup
  fit <- survfit(Surv(subdf[[time_col]], subdf[[status_col]]) ~ 1)
  s <- summary(fit)

  # event times and counts
  X  <- s$time                      # age/time at each event step
  nx <- s$n.risk                    # at risk just before the step
  dx <- s$n.event                   # deaths in the step

  # survivorship (relative to cohort start)
  l0 <- nx[1]
  lx <- nx / l0

  # conditional death probability over the step
  qx <- dx / nx

  # interval widths: from previous event time to current
  dt <- c(X[1], diff(X))
  dt[dt <= 0] <- NA_real_  # guard against zero/negative widths

  # force of mortality (piecewise-constant hazard)
  eps <- 1e-12
  qx_clamped <- pmin(pmax(qx, eps), 1 - eps)
  ux <- -log1p(-qx_clamped) / dt
  ln_ux <- log(ux)

  tibble(
    X      = X,
    nx     = nx,
    dx     = dx,
    lx     = lx,
    qx     = qx,
    ux     = ux,
    `ln(ux)` = ln_ux)
}

# 2 Build life tables for each sex
nd_data$Sex <- as.character(nd_data$Sex)
lifetables <- nd_data %>%
  group_by(Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$Sex))) %>%
  map(~ make_demography_lifetable(.x, time_col = "lifespan", status_col = "SAC"))

# 3 Optional: long-format table for plotting or export
lifetables_long <- lifetables %>%
  imap_dfr(~ mutate(.x, group = .y)) %>%
  separate(group, into = "Sex", sep = " ", remove = TRUE)

# 4 Make Gompertz model predictions

# Empty list to store hazard predictions
gompertz_preds <- list()

# Loop over each  sex
  for(s in unique(nd_data$Sex)) {
    # Subset data
    subdf <- nd_data %>% filter(Sex == s)
    
    # Fit Gompertz: flexsurvreg expects Surv(time, status)
    fit <- flexsurvreg(Surv(lifespan/365.25, SAC) ~ 1, data = subdf, dist = "gompertz")
    
    # Extract fitted parameters
    # "shape" = beta, "rate" = eta in flexsurv
    shape <- fit$res["shape","est"]
    rate  <- fit$res["rate","est"]
    
    # Prediction times
    times <- seq(0, max(subdf$lifespan/365.5, na.rm = TRUE), by = 1)
    
    # Compute hazard at each time
    hazard <- hgompertz(times, shape = shape, rate = rate)
    
    # Store in a data frame
    gompertz_preds[[paste(s, sep = "_")]] <- data.frame(
      X = times,
      hazard = hazard,
      Sex = s)}

# Combine all predictions into one long data frame
gompertz_preds_df <- bind_rows(gompertz_preds)


# 5 Age-specific mortality plot faceted by sex
mortality <- ggplot(lifetables_long, aes(x = X/365.25, y = ux, color = Sex)) +
  geom_point(size = 2, alpha = 0.5) +   # semi-transparent points
  geom_line(
    data = gompertz_preds_df,
    aes(x = X, y = hazard, color = Sex),
    size = 1.5,                         # thicker lines
    linetype = "solid") +                   # solid line for visibility
  # scale_color_manual(values = genotype_colors) +
  labs(
    x = "Age (years)",
    y = "Mortality rate",
    color = "Sex",
    title = "Age-specific mortality with Gompertz hazard") +
  ylim(0, .5) +
  theme_minimal(base_size = 14)
mortality

```
Ok, so this plots, but it's not correct, and I think it has to do with the way I encoded lifespan as years vs. days. I need to fix that. I am not sure years is the way to go, as that just gives me 7 time points to work with... Maybe I should try months, but that's still about 300 intervals, and I likely don't have enough data for that. Maybe every 6 months or every 3 months???  If I do 6 months, then I will have about 20 intervals.... That's 182.625 days, so I need to divide my lifespan by that number...

I will try that first, and then also go to include all data, not just the natural deaths. 

```{r}
# Remove the variables so I can reuse them without trouble.
rm(make_demography_lifetable, lifetables, gompertz_preds, gompertz_preds_df, mortality)

# 0 Recode Status to numeric (Dead = 1, others = 0)
# This first attempt only has natural death data.
summary(nd_data)
# Changing lifespan to 3 month intervals
nd_data2 <- nd_data
nd_data2$lifespan <- nd_data2$lifespan/91.3125

# 1 Function: actuarial/demography life table at event times
make_demography_lifetable <- function(subdf, time_col = "lifespan", status_col = "SAC") {
  # keep only rows with non-missing Death_Day/status
  subdf <- subdf %>% filter(!is.na(.data[[time_col]]), !is.na(.data[[status_col]])) %>%
    arrange(.data[[time_col]])
  
  # fit KM for this subgroup
  fit <- survfit(Surv(subdf[[time_col]], subdf[[status_col]]) ~ 1)
  s <- summary(fit)

  # event times and counts
  X  <- s$time                      # age/time at each event step
  nx <- s$n.risk                    # at risk just before the step
  dx <- s$n.event                   # deaths in the step

  # survivorship (relative to cohort start)
  l0 <- nx[1]
  lx <- nx / l0

  # conditional death probability over the step
  qx <- dx / nx

  # interval widths: from previous event time to current
  dt <- c(X[1], diff(X))
  dt[dt <= 0] <- NA_real_  # guard against zero/negative widths

  # force of mortality (piecewise-constant hazard)
  eps <- 1e-12
  qx_clamped <- pmin(pmax(qx, eps), 1 - eps)
  ux <- -log1p(-qx_clamped) / dt
  ln_ux <- log(ux)

  tibble(
    X      = X,
    nx     = nx,
    dx     = dx,
    lx     = lx,
    qx     = qx,
    ux     = ux,
    `ln(ux)` = ln_ux)
}

# 2 Build life tables for each sex
nd_data2$Sex <- as.character(nd_data2$Sex)
lifetables <- nd_data2 %>%
  group_by(Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$Sex))) %>%
  map(~ make_demography_lifetable(.x, time_col = "lifespan", status_col = "SAC"))

# 3 Optional: long-format table for plotting or export
lifetables_long <- lifetables %>%
  imap_dfr(~ mutate(.x, group = .y)) %>%
  separate(group, into = "Sex", sep = " ", remove = TRUE)

# 4 Make Gompertz model predictions

# Empty list to store hazard predictions
gompertz_preds <- list()

# Loop over each  sex
  for(s in unique(nd_data2$Sex)) {
    # Subset data
    subdf <- nd_data2 %>% filter(Sex == s)
    
    # Fit Gompertz: flexsurvreg expects Surv(time, status)
    fit <- flexsurvreg(Surv(lifespan, SAC) ~ 1, data = subdf, dist = "gompertz")
    
    # Extract fitted parameters
    # "shape" = beta, "rate" = eta in flexsurv
    shape <- fit$res["shape","est"]
    rate  <- fit$res["rate","est"]
    
    # Prediction times
    times <- seq(0, max(subdf$lifespan, na.rm = TRUE), by = 1)
    
    # Compute hazard at each time
    hazard <- hgompertz(times, shape = shape, rate = rate)
    
    # Store in a data frame
    gompertz_preds[[paste(s, sep = "_")]] <- data.frame(
      X = times,
      hazard = hazard,
      Sex = s)}

# Combine all predictions into one long data frame
gompertz_preds_df <- bind_rows(gompertz_preds)


# 5 Age-specific mortality plot faceted by sex
mortality <- ggplot(lifetables_long, aes(x = X, y = ux, color = Sex)) +
  geom_point(size = 2, alpha = 0.5) +   # semi-transparent points
  geom_line(
    data = gompertz_preds_df,
    aes(x = X, y = hazard, color = Sex),
    size = 1.5,                         # thicker lines
    linetype = "solid") +                   # solid line for visibility
  # scale_color_manual(values = genotype_colors) +
  labs(
    x = "Age (quarters)",
    y = "Mortality rate",
    color = "Sex",
    title = "Age-specific mortality with Gompertz hazard") +
  ylim(0, .5) +
  theme_minimal(base_size = 14)
mortality

```
This does look better. 

So, next, I will try this with all the data. 
```{r}
# Remove the variables so I can reuse them without trouble.
rm(make_demography_lifetable, lifetables, gompertz_preds, gompertz_preds_df, mortality)
rm(nd_data2)

# 0 Recode Status to numeric (Dead = 1, others = 0)

# Changing lifespan to 3 month intervals
nd_data2 <- data.fit
nd_data2$lifespan <- nd_data2$lifespan/91.3125

# 1 Function: actuarial/demography life table at event times
make_demography_lifetable <- function(subdf, time_col = "lifespan", status_col = "SAC") {
  # keep only rows with non-missing Death_Day/status
  subdf <- subdf %>% filter(!is.na(.data[[time_col]]), !is.na(.data[[status_col]])) %>%
    arrange(.data[[time_col]])
  
  # fit KM for this subgroup
  fit <- survfit(Surv(subdf[[time_col]], subdf[[status_col]]) ~ 1)
  s <- summary(fit)

  # event times and counts
  X  <- s$time                      # age/time at each event step
  nx <- s$n.risk                    # at risk just before the step
  dx <- s$n.event                   # deaths in the step

  # survivorship (relative to cohort start)
  l0 <- nx[1]
  lx <- nx / l0

  # conditional death probability over the step
  qx <- dx / nx

  # interval widths: from previous event time to current
  dt <- c(X[1], diff(X))
  dt[dt <= 0] <- NA_real_  # guard against zero/negative widths

  # force of mortality (piecewise-constant hazard)
  eps <- 1e-12
  qx_clamped <- pmin(pmax(qx, eps), 1 - eps)
  ux <- -log1p(-qx_clamped) / dt
  ln_ux <- log(ux)

  tibble(
    X      = X,
    nx     = nx,
    dx     = dx,
    lx     = lx,
    qx     = qx,
    ux     = ux,
    `ln(ux)` = ln_ux)
}

# 2 Build life tables for each sex
nd_data2$Sex <- as.character(nd_data2$Sex)
lifetables <- nd_data2 %>%
  group_by(Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$Sex))) %>%
  map(~ make_demography_lifetable(.x, time_col = "lifespan", status_col = "SAC"))

# 3 Optional: long-format table for plotting or export
lifetables_long <- lifetables %>%
  imap_dfr(~ mutate(.x, group = .y)) %>%
  separate(group, into = "Sex", sep = " ", remove = TRUE)

# 4 Make Gompertz model predictions

# Empty list to store hazard predictions
gompertz_preds <- list()

# Loop over each  sex
  for(s in unique(nd_data2$Sex)) {
    # Subset data
    subdf <- nd_data2 %>% filter(Sex == s)
    
    # Fit Gompertz: flexsurvreg expects Surv(time, status)
    fit <- flexsurvreg(Surv(lifespan, SAC) ~ 1, data = subdf, dist = "gompertz")
    
    # Extract fitted parameters
    # "shape" = beta, "rate" = eta in flexsurv
    shape <- fit$res["shape","est"]
    rate  <- fit$res["rate","est"]
    
    # Prediction times
    times <- seq(0, max(subdf$lifespan, na.rm = TRUE), by = 1)
    
    # Compute hazard at each time
    hazard <- hgompertz(times, shape = shape, rate = rate)
    
    # Store in a data frame
    gompertz_preds[[paste(s, sep = "_")]] <- data.frame(
      X = times,
      hazard = hazard,
      Sex = s)}

# Combine all predictions into one long data frame
gompertz_preds_df <- bind_rows(gompertz_preds)


# 5 Age-specific mortality plot faceted by sex
# I plotted the age in years, despite doing the analysis with quarters.
mortality <- ggplot(lifetables_long, aes(x = X/4, y = ux, color = Sex)) +
  geom_point(size = 2, alpha = 0.5) +   # semi-transparent points
  geom_line(
    data = gompertz_preds_df,
    aes(x = X/4, y = hazard, color = Sex),
    size = 1.5,                         # thicker lines
    linetype = "solid") +                   # solid line for visibility
  # scale_color_manual(values = genotype_colors) +
  labs(
    x = "Age (years)",
    y = "Mortality rate",
    color = "Sex",
    title = "Age-specific mortality with Gompertz hazard") +
  ylim(0, .5) +
  theme_minimal(base_size = 14)
mortality

# Exporting the file
pdf(file = "Mortality.pdf", width = 8, height = 5)

mortality

dev.off()

```


Next, I want to try and extract the mortality measures, shape and rate parameters from the model, and test for sex differences. 

```{r}
fit <- flexsurvreg(
  Surv(lifespan, SAC) ~ Sex,
  data = nd_data2,
  dist = "gompertz",
  anc = list(shape = ~ Sex)
)

coef_table <- fit$coefficients
vcov_mat <- vcov(fit)

library(tibble)
library(dplyr)

# Wald test
pvalues_table <- tibble(
  term = names(coef_table),
  estimate = coef_table,
  se = sqrt(diag(vcov_mat))
) %>%
  mutate(
    z = estimate / se,
    p_value = 2 * (1 - pnorm(abs(z)))
  )

pvalues_table <- pvalues_table %>%
  mutate(
    parameter = ifelse(grepl("^shape", term), "shape", "rate"),
    effect = gsub("shape:|Sex", "", term)
  )

pvalues_table
# "Aging rate (shape)", "Baseline hazard (rate)"
```


05/07/2026
After discussions with the co-authors, I need to try to run the mortality analysis differently - I need to actually create "bins" or time periods to record deaths in, so that I don't end up with unique time values, but instead intervals where, ideally, there is a death in each interval. So, I will again take the values that I had for lifespan, and divided them by 6 months or 182.625 days, but then, I will round this number up to the next integer to know which quarter the animals died in. And then, I will redo the analysis. 

## Prepping the data
```{r}
new.data <- data.fit %>% mutate(lifespan = ceiling(lifespan/(182.625)))
```

## Sanity check to make sure nothing weird happened with the survival analysis
```{r}
surv <- Surv(time = new.data$lifespan, event = new.data$SAC, type = "right")
fitsex <- survfit(formula= surv ~ Sex, data = new.data) # uses default Kaplan-Meier
summary(fitsex)
kap.sex<-ggsurvplot(
  fit = fitsex, 
  data = new.data,
  pval = TRUE, 
  surv.median.line = "hv",
  xlab = "Survival time (6 months)", 
  ylab = "Survival probability", 
  title = "Kaplan Meier's Plot by Sex"
  )
print(kap.sex)

# Exporting the file
pdf(file = "Kaplan_Meier_all_6_months.pdf", width = 8, height = 5)

kap.sex

dev.off()
```
Looks the same as before, sanity check complete. 


## Mortality again

So, next, I will try this with all the data. After some trial and error, I arrived at 6-month bins. 
```{r}
# Remove the variables so I can reuse them without trouble.
rm(make_demography_lifetable, lifetables, gompertz_preds, gompertz_preds_df, mortality)
rm(nd_data2)

# 0 Recode Status to numeric (Dead = 1, others = 0)

# Changing lifespan to 6 month intervals
nd_data2 <- data.fit
nd_data2$lifespan <- ceiling(nd_data2$lifespan/(91.3125*2))

# 1 Function: actuarial/demography life table at event times
make_demography_lifetable <- function(subdf, time_col = "lifespan", status_col = "SAC") {
  # keep only rows with non-missing Death_Day/status
  subdf <- subdf %>% filter(!is.na(.data[[time_col]]), !is.na(.data[[status_col]])) %>%
    arrange(.data[[time_col]])
  
  # fit KM for this subgroup
  fit <- survfit(Surv(subdf[[time_col]], subdf[[status_col]]) ~ 1)
  s <- summary(fit)

  # event times and counts
  X  <- s$time                      # age/time at each event step
  nx <- s$n.risk                    # at risk just before the step
  dx <- s$n.event                   # deaths in the step

  # survivorship (relative to cohort start)
  l0 <- nx[1]
  lx <- nx / l0

  # conditional death probability over the step
  qx <- dx / nx

  # interval widths: from previous event time to current
  dt <- c(X[1], diff(X))
  dt[dt <= 0] <- NA_real_  # guard against zero/negative widths

  # force of mortality (piecewise-constant hazard)
  eps <- 1e-12
  qx_clamped <- pmin(pmax(qx, eps), 1 - eps)
  ux <- -log1p(-qx_clamped) / dt
  ln_ux <- log(ux)

  tibble(
    X      = X,
    nx     = nx,
    dx     = dx,
    lx     = lx,
    qx     = qx,
    ux     = ux,
    `ln(ux)` = ln_ux)
}

# 2 Build life tables for each sex
nd_data2$Sex <- as.character(nd_data2$Sex)
lifetables <- nd_data2 %>%
  group_by(Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$Sex))) %>%
  map(~ make_demography_lifetable(.x, time_col = "lifespan", status_col = "SAC"))

# 3 Optional: long-format table for plotting or export
lifetables_long <- lifetables %>%
  imap_dfr(~ mutate(.x, group = .y)) %>%
  separate(group, into = "Sex", sep = " ", remove = TRUE)

# 4 Make Gompertz model predictions

# Empty list to store hazard predictions
gompertz_preds <- list()

# Loop over each  sex
  for(s in unique(nd_data2$Sex)) {
    # Subset data
    subdf <- nd_data2 %>% filter(Sex == s)
    
    # Fit Gompertz: flexsurvreg expects Surv(time, status)
    fit <- flexsurvreg(Surv(lifespan, SAC) ~ 1, data = subdf, dist = "gompertz")
    
    # Extract fitted parameters
    # "shape" = beta, "rate" = eta in flexsurv
    shape <- fit$res["shape","est"]
    rate  <- fit$res["rate","est"]
    
    # Prediction times
    times <- seq(0, max(subdf$lifespan, na.rm = TRUE), by = 1)
    
    # Compute hazard at each time
    hazard <- hgompertz(times, shape = shape, rate = rate)
    
    # Store in a data frame
    gompertz_preds[[paste(s, sep = "_")]] <- data.frame(
      X = times,
      hazard = hazard,
      Sex = s)}

# Combine all predictions into one long data frame
gompertz_preds_df <- bind_rows(gompertz_preds)


# 5 Age-specific mortality plot faceted by sex
# I plotted the age in years, despite doing the analysis with quarters.
mortality <- ggplot(lifetables_long, aes(x = X/2, y = ux, color = Sex)) +
  geom_point(size = 2, alpha = 0.5) +   # semi-transparent points
  geom_line(
    data = gompertz_preds_df,
    aes(x = X/2, y = hazard, color = Sex),
    size = 1.5,                         # thicker lines
    linetype = "solid") +                   # solid line for visibility
  # scale_color_manual(values = genotype_colors) +
  labs(
    x = "Age (years)",
    y = "Mortality rate",
    color = "Sex",
    title = "Age-specific mortality with Gompertz hazard") +
  ylim(0, .5) +
  theme_minimal(base_size = 14)
mortality

# Exporting the file
pdf(file = "Mortality_semi.pdf", width = 8, height = 5)

mortality

dev.off()

```

This looks good, and based on the information below, I have deaths in each bin other than the last one. 

```{r}
nd_data2 %>%
     filter (!is.na(lifespan)) %>%
     mutate(bin = cut(lifespan, breaks = 21)) %>%
     group_by(bin) %>%
     summarise(count = n()) %>%
     print(n=21)
```

Next step is to actually compare the model parameters. 

Next, I want to try and extract the mortality measures, shape and rate parameters from the model, and test for sex differences. 

```{r}
rm(fit, coef_table, vcov_mat, pvalues_table)

fit <- flexsurvreg(
  Surv(lifespan, SAC) ~ Sex,
  data = nd_data2,
  dist = "gompertz",
  anc = list(shape = ~ Sex)
)

coef_table <- fit$coefficients
vcov_mat <- vcov(fit)

library(tibble)
library(dplyr)

# Wald test
pvalues_table <- tibble(
  term = names(coef_table),
  estimate = coef_table,
  se = sqrt(diag(vcov_mat))
) %>%
  mutate(
    z = estimate / se,
    p_value = 2 * (1 - pnorm(abs(z)))
  )

pvalues_table <- pvalues_table %>%
  mutate(
    parameter = ifelse(grepl("^shape", term), "shape", "rate"),
    effect = gsub("shape:|Sex", "", term)
  )

pvalues_table
# "Aging rate (shape)", "Baseline hazard (rate)"
```

## Getting some additional information for plots
In line 553, I made a survfit object called fitsex that contains the fitted Kaplan-Meier data for both sexes. I want to extract the survival table from that object for each sex. 

```{r}
summary(fitsex) # To check if the object looks ok

table_fitsex <- surv_summary(fitsex)

# Export to xlsx

write_xlsx(table_fitsex, "Tree_shrew_fit.xlsx")


```

A couple more basic stats for the manuscript
```{r}
# Summary stats for the lifespan analysis
# Removes anyone without lifespan
y <- step.4 %>% filter (!is.na(lifespan))
y[y[,8] == "L",8] <-"Living"
y[y[,8] == "Y",8] <-"Euthanized"
y[y[,8] == "N",8] <-"Natural Death"
y[y[,8] == "",8] <-"Unknown"
table(y$Sex,y$SAC)
```
y has the same rows than data.fit, but it retains the information about deaths, rather than just classifying them as censored or not. 

###################################################################################################
## Final analysis for manuscript
###################################################################################################

## Lifespan and mortality analysis excluding immature animals - final analysis used in manuscript
I talked to Anne, and she suggested that it is best practice to exclude animals that are not sexually mature yet. So, I need to exlcude animals that are less than 4-5 months old. Given that I start with a 6 months bin, I should exclude the first bin. I recoded lifespan to start at 6 months, which I will need to fix when making final plots etc. 

```{r}
# Make the new data frame
# Filtering out deaths from the first 6 months; setting 0 to 6-months for modeling
rm(surv, fitsex, kap.sex)
new.data2 <- new.data %>% filter (lifespan >1) %>% mutate(lifespan = lifespan-1)


surv <- Surv(time = new.data2$lifespan, event = new.data2$SAC, type = "right")
fitsex <- survfit(formula= surv ~ Sex, data = new.data2) # uses default Kaplan-Meier
summary(fitsex)
kap.sex<-ggsurvplot(
  fit = fitsex, 
  data = new.data2,
  pval = TRUE, 
  surv.median.line = "hv",
  xlab = "Survival time (6 months)", 
  ylab = "Survival probability", 
  title = "Kaplan Meier's Plot by Sex"
  )
print(kap.sex)

rm(surv_times)
surv_times <- quantile(fitsex, probs = quantile_probs)
surv_times

# Exporting the file
pdf(file = "Kaplan_Meier_all_6_months_mature.pdf", width = 8, height = 5)

kap.sex

dev.off()

# Comparing the lifespan curves
log_rank_test <- survdiff(surv ~ Sex, data = new.data2)
log_rank_test

# Comparing median lifespan
wilcox.test(lifespan ~ Sex, alternative = "greater", data = new.data2)


```
## Getting some additional information for plots
In line 1418, I made a survfit object called fitsex that contains the fitted Kaplan-Meier data for both sexes. I want to extract the survival table from that object for each sex. 

```{r}
summary(fitsex) # To check if the object looks ok

table_fitsex <- surv_summary(fitsex)

# Export to xlsx

write_xlsx(table_fitsex, "Tree_shrew_fit.xlsx")


```


# Looking at maximum lifespan differences
Fisher's test for maximum lifespan (Wang 2004)
```{r}
# compute 90th percentile cutoff from pooled distribution
cutoff <- quantile(new.data2$lifespan, 0.9, na.rm = TRUE)


  # classify long-lived vs not
  df <- new.data2 %>%
    mutate(
      top10 = ifelse(lifespan >= cutoff, "yes", "no"),
      group = Sex
    )

  # build contingency table
  tab <- table(df$group, df$top10)

  # run fisher test
  ft <- fisher.test(tab)
  ft

```

And now using these same filtered, mature animal only data for mortality analysis.
 
```{r}
# Remove the variables so I can reuse them without trouble.
rm(make_demography_lifetable, lifetables, gompertz_preds, gompertz_preds_df, mortality)
rm(nd_data2)

# 0 Recode Status to numeric (Dead = 1, others = 0)

# 1 Function: actuarial/demography life table at event times
make_demography_lifetable <- function(subdf, time_col = "lifespan", status_col = "SAC") {
  # keep only rows with non-missing Death_Day/status
  subdf <- subdf %>% filter(!is.na(.data[[time_col]]), !is.na(.data[[status_col]])) %>%
    arrange(.data[[time_col]])
  
  # fit KM for this subgroup
  fit <- survfit(Surv(subdf[[time_col]], subdf[[status_col]]) ~ 1)
  s <- summary(fit)

  # event times and counts
  X  <- s$time                      # age/time at each event step
  nx <- s$n.risk                    # at risk just before the step
  dx <- s$n.event                   # deaths in the step

  # survivorship (relative to cohort start)
  l0 <- nx[1]
  lx <- nx / l0

  # conditional death probability over the step
  qx <- dx / nx

  # interval widths: from previous event time to current
  dt <- c(X[1], diff(X))
  dt[dt <= 0] <- NA_real_  # guard against zero/negative widths

  # force of mortality (piecewise-constant hazard)
  eps <- 1e-12
  qx_clamped <- pmin(pmax(qx, eps), 1 - eps)
  ux <- -log1p(-qx_clamped) / dt
  ln_ux <- log(ux)

  tibble(
    X      = X,
    nx     = nx,
    dx     = dx,
    lx     = lx,
    qx     = qx,
    ux     = ux,
    `ln(ux)` = ln_ux)
}

# 2 Build life tables for each sex
new.data2$Sex <- as.character(new.data2$Sex)
lifetables <- new.data2 %>%
  group_by(Sex) %>%
  group_split() %>%
  set_names(map_chr(., ~ unique(.x$Sex))) %>%
  map(~ make_demography_lifetable(.x, time_col = "lifespan", status_col = "SAC"))

# 3 Optional: long-format table for plotting or export
lifetables_long <- lifetables %>%
  imap_dfr(~ mutate(.x, group = .y)) %>%
  separate(group, into = "Sex", sep = " ", remove = TRUE)

# 4 Make Gompertz model predictions

# Empty list to store hazard predictions
gompertz_preds <- list()

# Loop over each  sex
  for(s in unique(new.data2$Sex)) {
    # Subset data
    subdf <- new.data2 %>% filter(Sex == s)
    
    # Fit Gompertz: flexsurvreg expects Surv(time, status)
    fit <- flexsurvreg(Surv(lifespan, SAC) ~ 1, data = subdf, dist = "gompertz")
    
    # Extract fitted parameters
    # "shape" = beta, "rate" = eta in flexsurv
    shape <- fit$res["shape","est"]
    rate  <- fit$res["rate","est"]
    
    # Prediction times
    times <- seq(0, max(subdf$lifespan, na.rm = TRUE), by = 1)
    
    # Compute hazard at each time
    hazard <- hgompertz(times, shape = shape, rate = rate)
    
    # Store in a data frame
    gompertz_preds[[paste(s, sep = "_")]] <- data.frame(
      X = times,
      hazard = hazard,
      Sex = s)}

# Combine all predictions into one long data frame
gompertz_preds_df <- bind_rows(gompertz_preds)


# 5 Age-specific mortality plot faceted by sex
# I plotted the age in years, despite doing the analysis with 6-month periods
mortality <- ggplot(lifetables_long, aes(x = X/2, y = ux, color = Sex)) +
  geom_point(size = 2, alpha = 0.5) +   # semi-transparent points
  geom_line(
    data = gompertz_preds_df,
    aes(x = X/2, y = hazard, color = Sex),
    size = 1.5,                         # thicker lines
    linetype = "solid") +                   # solid line for visibility
  # scale_color_manual(values = genotype_colors) +
  labs(
    x = "Age (years)",
    y = "Mortality rate",
    color = "Sex",
    title = "Age-specific mortality with Gompertz hazard") +
  ylim(0, .5) +
  theme_minimal(base_size = 14)
mortality

# Exporting the file
pdf(file = "Mortality_semi_mature.pdf", width = 8, height = 5)

mortality

dev.off()

```
```{r}
rm(fit, coef_table, vcov_mat, pvalues_table)

fit <- flexsurvreg(
  Surv(lifespan, SAC) ~ Sex,
  data = new.data2,
  dist = "gompertz",
  anc = list(shape = ~ Sex)
)

coef_table <- fit$coefficients
vcov_mat <- vcov(fit)

library(tibble)
library(dplyr)

# Wald test
pvalues_table <- tibble(
  term = names(coef_table),
  estimate = coef_table,
  se = sqrt(diag(vcov_mat))
) %>%
  mutate(
    z = estimate / se,
    p_value = 2 * (1 - pnorm(abs(z)))
  )

pvalues_table <- pvalues_table %>%
  mutate(
    parameter = ifelse(grepl("^shape", term), "shape", "rate"),
    effect = gsub("shape:|Sex", "", term)
  )

pvalues_table
# "Aging rate (shape)", "Baseline hazard (rate)"

```

